---
title: "README"
author: "Kai Jiang"
output: github_document
---

# BSRMM
Bayesian Reduced Rank Regression for Microbiome-Metabolite Data Integration

**Author**: Kai Jiang  
**Reference**: Jiang K, Saha S, Peterson CB. Bayesian sparse regression for microbiome-metabolite data integration.

## Overview

This repository provides the implementation of **Bayesian Sparse Regression for Microbiome-Metabolite Data Integration (BSRMM)**. This method enables joint modeling of high-dimensional microbiome predictors and metabolomic outcomes. The BSRMM model is implemented in R and is designed for variable selection and missing value imputation for outcomes.

## Directory Structure

- `code/` – Contains scripts to generate simulated datasets and run the BSRMM model.
- `data/` – Contains simulated datasets.
- `function` - Contains the implementation of the BSRMM model and simulation functions.
- `plot` - Contains MCMC diagnostic plots.
- `result` - Contains the results of the BSRMM model and diagnostic R-hat values.

## Example

```{r, eval=TRUE, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
setwd("/Users/kjiang/Desktop/BSRMM/")

# functions 
source("./function/bsrmmgibbs.R")
source("./function/bsrmmbf.R")

# required library
library(truncnorm)
library(mvnfast)
library(MASS)
library(coda)
library(MCMCpack)
library(reshape2)
library(knitr)
library(ggplot2)
```

### Import simulated datasets
```{r, eval=TRUE, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
# import data 
mydata <- readRDS("./data/low_mnar0.33_ind_300_1000_1_0.3_1.RDS")

# extract predictors and outcomes
predictor <- mydata$X
outcome <- mydata$Y_miss
```

### Check the true nonzero coefficients
```{r comment=""}
cat("The nonzero coefficients are: ", which(mydata$coeffs[,1] != 0), "\n")
cat("The value of the nonzero coefficients are: ", mydata$coeffs[which(mydata$coeffs[,1] != 0),2], "\n")
```

### Split the data into train and test sets, and prepare for BSRMM model fitting
```{r}
# consider scale and center the predictor before split
stand <- list()
stand$mux <- colMeans(predictor)
stand$Sx <- apply(predictor, 2, sd)
predictor <- apply(predictor, 2, function(x) (x - mean(x))/sd(x))

# we already performed mean or low missing value imputation during that simulation generation
# we need to re-set the value into zero
index <- which(mydata$Y_miss != mydata$Y_true)
index <- sort(index)
outcome[index] <- 0

# save dt for following prediction error analysis: 0 = miss, 1 = observed
dt <- data.frame(outcome)
dt$Ri <- 0
dt$Ri[which(dt$outcome != 0)] <- 1

# split the data into train and test
set.seed(33)
seed <- 33

train_ind <- sample(nrow(predictor),size = round(0.7*nrow(predictor)),replace = FALSE)
train_ind <- sort(train_ind)
# train_ind

predictor_train <- predictor[train_ind,]
predictor_test <- predictor[-train_ind,]

outcome_train <- outcome[train_ind,]
outcome_test <- outcome[-train_ind,]

dt_train <- dt[train_ind,]
dt_test <- dt[-train_ind,]

X <- as.matrix(predictor_train)

# check the outcome 
Y <- as.matrix(outcome_train)

# preparation for gibbs sampling
n <- nrow(X)
p <- ncol(X)
c <- 100 # larger c to maintain the compositionality of the predictors 

# construct the T matrix, because we cannot assign a matrix to T due to R's limitation 
# we name it as N matrix 
N <- rbind(diag(x = 1, nrow = p, ncol = p), matrix(data = c, nrow = 1, ncol = p))

# number of sampling procedure 
nburnin <- 10000
niter <- 20000

# initial number of predictors 
nop <- floor(n/2)
```

### Construct the Q matrix and initialize the hyperparameters for BSRMM model fitting
```{r, cache=TRUE}
Q <- diag(0,nrow = p) 

# initial hyperparameters 
tau <- 1 
nu <- 0
omega <- 0 

a0 <- -12
a <- rep(a0,p)

predict_result = TRUE
display = FALSE
bayestmetab = TRUE

# save the running time: 
bsrmm <- bsrmmgibbs(nburnin, niter, p, nop, Y, X, N, a, Q, n, tau, nu, omega, 
                    alpha_a_shape = 1, alpha_b_shape = 1, seed, predict_result, bayestmetab, display, lod_ratio = 1)

# Result
PPI <- colMeans(bsrmm$gamma[(nburnin+2):(nburnin+niter+1),]) # posterior probabilities of inclusion
beta_select <- rep(0, ncol(X))
beta_select[which(PPI > 0.5)] <- 1
```

### Check selected predictors vs true nonzero coefficients
```{r, comment=""}
selected <- which(PPI > 0.5)
true_nz  <- which(mydata$coeffs[,1] != 0)

cat(sprintf("Selection Results:\n"))
cat(sprintf(" - Selected Predictors: %s\n", paste(selected, collapse = ", ")))
cat(sprintf(" - True Predictors:     %s\n", paste(true_nz, collapse = ", ")))
cat(sprintf(" - Accuracy: %d out of %d true signals recovered.\n", 
            length(intersect(selected, true_nz)), length(true_nz)))
```

### Coefficients estimation plots
```{r}
beta_esti <- bsrmm$tembeta_save
beta_esti <- beta_esti[which(PPI > 0.5), (nburnin+2):(nburnin+niter+1)]
beta_esti[is.na(beta_esti)] <- 0
beta_esti <- sweep(beta_esti, 1, STATS =  stand$Sx[which(PPI > 0.5)], FUN = "/")
beta_result <- data.frame(
  param = c(paste0("beta", which(PPI > 0.5))),
  mean = rowMeans(beta_esti),
  Lower_CrI = apply(beta_esti, 1, function(x) quantile(x, probs = 0.025)),
  Upper_CrI = apply(beta_esti, 1, function(x) quantile(x, probs = 0.975))
)

beta_result$True_value <- mydata$coeffs[which(PPI > 0.5), 2]

par(mfrow =c(2,3))

for (i in 1:nrow(beta_result)) {
  # Plot the trace of the MCMC samples
  plot(beta_esti[i, ], 
       type = "l", 
       main = paste0("Coefficient: ", beta_result$param[i]), 
       xlab = "MCMC Iteration", 
       ylab = "Value",
       col  = "steelblue", 
       lwd  = 1)
  
  # Add Ground Truth (Dashed Red)
  abline(h = beta_result$True_value[i], col = "red", lwd = 2, lty = 2)
}
```

### Missing value imputation performance

```{r}
# posterior mean for MVI
Y_mvi <- bsrmm$Y_mvi[(which(bsrmm$Ri==0)), (nburnin+2):(nburnin+niter+1)]
Y_true <- as.matrix(mydata$Y_true)
Y_true_train <- as.matrix(Y_true[train_ind, ])
Y_true_train <- Y_true_train[which(bsrmm$Ri==0), ]

mvi_result <- rowMeans(Y_mvi)
mvi_result <- data.frame(
  ground_truth = Y_true_train,
  mean = rowMeans(Y_mvi),
  Lower_CrI = apply(Y_mvi, 1, function(x) quantile(x, probs = 0.025)),
  Upper_CrI = apply(Y_mvi, 1, function(x) quantile(x, probs = 0.975))
)

ggplot(mvi_result, aes(x = ground_truth, y = mean)) +
    # Add the vertical error bars for CrI
    geom_errorbar(aes(ymin = Lower_CrI, ymax = Upper_CrI), 
                  width = 0.1, color = "gray", alpha = 0.6) +
    # Add the scatter points
    geom_point(color = "blue", size = 1.5) +
    # Add a 45-degree reference line (Ideal prediction where mean == ground_truth)
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
    # Labels and Theme
    labs(x = "Ground truth values",
         y = "Imputed values from BSRMM with 95% CrI") +
    theme_classic()
  

```

